---
output: html_document
---
<p align="right">Bennett Widyolar<br>
ES 207 - HW #7<br>
4/7/2016</p>

<center>
# Homework Assignment 7
</center>

## Objective
The objective of this study is to determine the percentage of years it floods more than 100 days (floods defined as discharges >= 800 cubic feet per second) along the Cosumnes River, to assist the California Department of Water Resources in determining whether or not setback levee construction is a good use of Prop. 1 funds for the purpose of groundwater recharge.

## Methods
In this study, we will analyze discharge measurements along the Cosumnes River to determine the number of flood events per year. The percentage of years having greater than 100 flood days will be reported back to the DWR.

## Data
The data used has been collected by the USGS since 1907 for the following site:
Site #: 11335000
Site Name: Michigan Bar
Sensor: Discharge, cubic feet per second (mean)

It is available at the following url: http://waterdata.usgs.gov/nwis/dv?cb_00060=on&format=rdb&site_no=11335000&referred_module=sw&period=&begin_date=1907-10-01&end_date=2015-09-30

## Code
The code for this project is as follows:

```{r, eval=FALSE}
#CLEAR VARIABLES
rm(list=ls())

site='11335000'
sensor='00060'
start='1907-10-01'
end='2015-09-30'

data <- read.table(paste("http://waterdata.usgs.gov/nwis/dv?cb_",sensor,"=on&format=rdb&site_no=", site,"&referred_module=sw&period=&begin_date=",start,"&end_date=", end, sep=""),
                   header=T, sep="", quote="'", skip=0, colClasses = c("factor", "factor", "factor", "factor", "factor"))
print('DONE')

#REMOVE JUNK DATA FROM FIRST ROW
data = droplevels(data[-1,])
#RENAME LAST TWO COLUMNS
colnames(data)[4:5]= c("discharge_cfs","dat_qvalue")

data$datetime = as.POSIXct(data$datetime, "America/Los_Angeles")
data$discharge_cfs = as.double(as.character(data$discharge_cfs))

data$discharge_afd = data$discharge_cfs*1.98347
#NOTE POSSIBLE OUTLIER - 122,200 AF/Day?

#WATER YEAR
data$year = as.numeric(substr(data$datetime,1,4))
data$month = as.numeric(substr(data$datetime,6,7))
data$day = as.numeric(substr(data$datetime,9,11))
data$wy = ifelse(data$month > 9, data$year +1, data$year)
data$wym = ifelse(data$month > 9, data$month - 9, data$month + 3)

#REMOVE LEAP YEAR
data <- data[!(data$month ==2 & data$day == 29),]

#TOTAL DISCHARGE BY MONTH
data.mo.afd.sum = ts(as.vector(tapply(data$discharge_afd,list(data$wym,data$wy),sum)),frequency=12)
data.mo.afd.mean = ts(as.vector(tapply(data$discharge_afd,list(data$wym,data$wy),mean)),frequency=12)
data.mo.afd.max = ts(as.vector(tapply(data$discharge_afd,list(data$wym,data$wy),max)),frequency=12)
data.mo.cfs.mean = ts(as.vector(tapply(data$discharge_cfs,list(data$wym,data$wy),mean)),frequency=12)
data.mo.cfs.max = ts(as.vector(tapply(data$discharge_cfs,list(data$wym,data$wy),max)),frequency=12)

data.yr.afd.sum = ts(as.vector(tapply(data$discharge_afd,list(data$wy),sum)),frequency=1)
data.yr.afd.mean = ts(as.vector(tapply(data$discharge_afd,list(data$wy),mean)),frequency=1)

#DRIEST YEAR
index_min = which.min(data.yr.afd.sum)
plot(data.mo.afd.mean[(index_min*12+1):(index_min*12+12)], type="o")
#WETTEST YEAR
index_max = which.max(data.yr.afd.sum)
plot(data.mo.afd.mean[(index_max*12+1):(index_max*12+12)], type="o")
#HIGHEST DAILY AVERAGE DISCHARGE
index_discharge = which.max(data.yr.afd.mean)
plot(data.mo.afd.mean[(index_discharge*12+1):(index_discharge*12+12)], type="o")

plot(stl(data.mo.afd.mean,s.window="periodic"))
plot(acf(data.mo.afd.mean))

#FLOODS
plot(data$discharge_cfs)
threshold = 800
abline(h=threshold, col="red")

data$flood <- data$discharge_cfs >= threshold
data.yr.flood.sum <- ts(as.vector(tapply(data$flood, list(data$wy),sum)), start=1907)
print(sum(data.yr.flood.sum >= 100)/length(data.yr.flood.sum))
```

## Results

```{r, echo=FALSE}
#CLEAR VARIABLES
rm(list=ls())

site='11335000'
sensor='00060'
start='1907-10-01'
end='2015-09-30'

data <- read.table(paste("http://waterdata.usgs.gov/nwis/dv?cb_",sensor,"=on&format=rdb&site_no=", site,"&referred_module=sw&period=&begin_date=",start,"&end_date=", end, sep=""),
                   header=T, sep="", quote="'", skip=0, colClasses = c("factor", "factor", "factor", "factor", "factor"))
print('DONE')

#REMOVE JUNK DATA FROM FIRST ROW
data = droplevels(data[-1,])
#RENAME LAST TWO COLUMNS
colnames(data)[4:5]= c("discharge_cfs","dat_qvalue")

data$datetime = as.POSIXct(data$datetime, "America/Los_Angeles")
data$discharge_cfs = as.double(as.character(data$discharge_cfs))

data$discharge_afd = data$discharge_cfs*1.98347
#NOTE POSSIBLE OUTLIER - 122,200 AF/Day?

#WATER YEAR
data$year = as.numeric(substr(data$datetime,1,4))
data$month = as.numeric(substr(data$datetime,6,7))
data$day = as.numeric(substr(data$datetime,9,11))
data$wy = ifelse(data$month > 9, data$year +1, data$year)
data$wym = ifelse(data$month > 9, data$month - 9, data$month + 3)

#REMOVE LEAP YEAR
data <- data[!(data$month ==2 & data$day == 29),]

#TOTAL DISCHARGE BY MONTH
data.mo.afd.sum = ts(as.vector(tapply(data$discharge_afd,list(data$wym,data$wy),sum)),frequency=12)
data.mo.afd.mean = ts(as.vector(tapply(data$discharge_afd,list(data$wym,data$wy),mean)),frequency=12)
data.mo.afd.max = ts(as.vector(tapply(data$discharge_afd,list(data$wym,data$wy),max)),frequency=12)
data.mo.cfs.mean = ts(as.vector(tapply(data$discharge_cfs,list(data$wym,data$wy),mean)),frequency=12)
data.mo.cfs.max = ts(as.vector(tapply(data$discharge_cfs,list(data$wym,data$wy),max)),frequency=12)

data.yr.afd.sum = ts(as.vector(tapply(data$discharge_afd,list(data$wy),sum)),frequency=1)
data.yr.afd.mean = ts(as.vector(tapply(data$discharge_afd,list(data$wy),mean)),frequency=1)

#DRIEST YEAR
index_min = which.min(data.yr.afd.sum)
plot(data.mo.afd.mean[(index_min*12+1):(index_min*12+12)], type="o", main="Driest Year Hydrograph - 1976")
#WETTEST YEAR
index_max = which.max(data.yr.afd.sum)
plot(data.mo.afd.mean[(index_max*12+1):(index_max*12+12)], type="o", main="Wettest Year Hydrograph - 1982")
#HIGHEST DAILY AVERAGE DISCHARGE
index_discharge = which.max(data.yr.afd.mean)
plot(data.mo.afd.mean[(index_discharge*12+1):(index_discharge*12+12)], type="o", main="Hydrograph of year with highest daily average discharge - 1982")
```

The following charts depict the temporal autocorrelation in the monthly discharge.
There is clearly a seasonal pattern (sine wave), with high discharge in the spring and summer and low discharge during the fall and winter, that repeats on a yearly basis.

```{r, echo=FALSE}
plot(stl(data.mo.afd.mean,s.window="periodic"))
plot(acf(data.mo.afd.mean))
```

Flood analysis is performed below. The hydrograph (Daily) since 1907 is plotted along with the drawn threshold of 800 cfs at which a day is counted as a flood event.

```{r, echo=FALSE}
#FLOODS
plot(data$discharge_cfs)
threshold = 800
abline(h=threshold, col="red")
```

The percentage of years from 1907-2015 with more than 100 flood events is evaluated below.
```{r,echo=FALSE}
data$flood <- data$discharge_cfs >= threshold
data.yr.flood.sum <- ts(as.vector(tapply(data$flood, list(data$wy),sum)), start=1907)
print(sum(data.yr.flood.sum >= 100)/length(data.yr.flood.sum))
```

## Discussion

The results show that slightly less than 1/3 of all years flood for more than 100 days. Thus, based on the > 1/3 of years, setback levee construction will probably not be pursued under Prop 1. funds.

## Limitations
See header of data file: http://waterdata.usgs.gov/nwis/dv?cb_00060=on&format=rdb&site_no=11335000&referred_module=sw&period=&begin_date=1907-10-01&end_date=2015-09-30