---
output: html_document
---
<p align="right">Bennett Widyolar<br>
ES 207 - HW #8<br>
4/13/2016</p>

<center>
# Homework Assignment 8
</center>

## Objective
The objective of this study is to provide followup information towards determining whether setback levees are a good use of Prop. 1 funds by assessing whether or not the water quality of floodplains is conducive to fish growth.

## Methods
In this study we analyze the quality of water at the Triangle Floodplain, just north of the Accidental Forest in the Cosumnes River Preserve.

## Data
The data used for this analysis came from the shape file: allobs.shp, downloaded from this course website.

## Code
The code for this project is as follows:

```{r, eval=FALSE}
#CLEAR VARIABLES
rm(list=ls())
#RESET PLOT
par(mfrow=c(1,1))

#install.packages("maptools")
#install(akima)
library(maptools)
library(akima)

pt.shp <- readShapePoints("ahearn_allobs/allobs")

pt.shp.18 <- subset(pt.shp, DateTime_ == "2005-02-18")
pt.shp.23 <- subset(pt.shp, DateTime_ == "2005-02-23")

pt.shp.18.temp <- interp(pt.shp.18$X, pt.shp.18$Y, pt.shp.18$Temp, duplicate="mean")
pt.shp.18.chlo <- interp(pt.shp.18$X, pt.shp.18$Y, pt.shp.18$Chlorophyl, duplicate="mean")
pt.shp.18.do <- interp(pt.shp.18$X, pt.shp.18$Y, pt.shp.18$DO__, duplicate="mean")

par(mfrow=c(2,3))

image(pt.shp.18.temp, col=cm.colors(24))
contour(pt.shp.18.temp, add=T)
image(pt.shp.18.chlo, col=topo.colors(24))
contour(pt.shp.18.chlo, add=T)
image(pt.shp.18.do, col=heat.colors(24))
contour(pt.shp.18.do, add=T)

persp(pt.shp.18.temp, theta=-50, phi=45, col="lightblue",zlab="Temp C", xlab="UTM X (m)", ylab="UTM Y (m)")
persp(pt.shp.18.chlo, theta=-50, phi=45, col="lightblue",zlab="Chlorophyll mg/L", xlab="UTM X (m)", ylab="UTM Y (m)")
persp(pt.shp.18.do, theta=-50, phi=45, col="lightblue",zlab="Dissolved Oxygen mg/L", xlab="UTM X (m)", ylab="UTM Y (m)")


```

## Results

The following charts plot the Temperature, Chlorophyll concentration, and Dissolved Oxygen measurements in the Triangle Floodplain.

```{r, echo=F}
#CLEAR VARIABLES
rm(list=ls())
#RESET PLOT
par(mfrow=c(1,1))

#install.packages("maptools")
#install(akima)
library(maptools)
library(akima)

pt.shp <- readShapePoints("ahearn_allobs/allobs")

pt.shp.18 <- subset(pt.shp, DateTime_ == "2005-02-18")
pt.shp.23 <- subset(pt.shp, DateTime_ == "2005-02-23")

pt.shp.18.temp <- interp(pt.shp.18$X, pt.shp.18$Y, pt.shp.18$Temp, duplicate="mean")
pt.shp.18.chlo <- interp(pt.shp.18$X, pt.shp.18$Y, pt.shp.18$Chlorophyl, duplicate="mean")
pt.shp.18.do <- interp(pt.shp.18$X, pt.shp.18$Y, pt.shp.18$DO__, duplicate="mean")

par(mfrow=c(2,3))

image(pt.shp.18.temp, col=cm.colors(24))
contour(pt.shp.18.temp, add=T)
image(pt.shp.18.chlo, col=topo.colors(24))
contour(pt.shp.18.chlo, add=T)
image(pt.shp.18.do, col=heat.colors(24))
contour(pt.shp.18.do, add=T)

persp(pt.shp.18.temp, theta=-50, phi=45, col="lightblue",zlab="Temp C", xlab="UTM X (m)", ylab="UTM Y (m)")
persp(pt.shp.18.chlo, theta=-50, phi=45, col="lightblue",zlab="Chlorophyll mg/L", xlab="UTM X (m)", ylab="UTM Y (m)")
persp(pt.shp.18.do, theta=-50, phi=45, col="lightblue",zlab="Dissolved Oxygen mg/L", xlab="UTM X (m)", ylab="UTM Y (m)")


```

Additionally, an interactive Shiny App is developed and allows the user to view water quality information for different days (see Shiny App files: ui.R and server.R)

## Limitations
Duplicate points in the shapefile are averaged.